---
title: "Prediction Assignment"
author: "NBP"
date: "Friday, January 15, 2016"
output: html_document
---

##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

##Prepare datasets

Get training data into data table.

```{r}
require(data.table)
setInternet2(TRUE)
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
DTrain <- fread(url)
```
Get test data into data table.
 
```{r}
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
DTest <- fread(url)
```

Only column names containing belt, arm, dumbell and forearms that has no missing values in the training dataset will be **predictor**.

```{r}
isAnyMissing <- sapply(DTrain, function (x) any(is.na(x) | x == ""))
isPredictor <- !isAnyMissing & grepl("belt|[^(fore)]arm|dumbbell|forearm", names(isAnyMissing))
predCandidates <- names(isAnyMissing)[isPredictor]
predCandidates
```

Subset the primary dataset to include only the **predictor** and the outcome varibale, **classe**.

```{r}
varToInclude <- c("classe", predCandidates)
DTrain <- DTrain[, varToInclude, with=FALSE]
dim(DTrain)
names(DTrain)
```

Make classe into a factor.

```{r}
DTrain <- DTrain[, classe := factor(DTrain[, classe])]
DTrain[, .N, classe]
```

Split the dataset into 60% training and 40% probing dataset.

```{r}
require(caret)
set.seed(12345)
inTrain <- createDataPartition(DTrain$classe, p=0.6)
DTraining <- DTrain[inTrain[[1]]]
DProbing <- DTrain[-inTrain[[1]]]
```

Preprocess the prediction variables by centering and scaling.

```{r}
X <- DTraining[, predCandidates, with=FALSE]
preProc <- preProcess(X)
preProc
XCS <- predict(preProc, X)
DTrainingCS <- data.table(data.frame(classe = DTraining[, classe], XCS))
```

Apply the centering and scaling to the probing dataset.

```{r}
X <- DProbing[, predCandidates, with=FALSE]
XCS <- predict(preProc, X)
DProbingCS <- data.table(data.frame(classe = DProbing[, classe], XCS))
```

Check for near zero variance

```{r}
nzv <- nearZeroVar(DTrainingCS, saveMetrics=TRUE)
if(any(nzv$nzv)) nzv else message ("No variables with near zero variance")
```

Examine groups of prediction variables.

```{r}
histGroup <- function (data, regex) {
  col <- grep(regex, names(data))
  col <- c(col, which(names(data) == "classe"))
  require(reshape2)
  n <- nrow(data)
  DMelted <- melt(data[, col, with=FALSE][, rownum := seq(1, n)], id.vars=c("rownum", "classe"))
  require(ggplot2)
  ggplot(DMelted, aes(x=classe, y=value)) +
    geom_violin(aes(color=classe, fill=classe), alpha=1/2) +
    facet_wrap(~ variable, scale="free_y") +
    scale_color_brewer(palette="Spectral") +
    scale_fill_brewer(palette="Spectral") +
    labs(x="", y="") +
    theme(legend.position="none")
}
histGroup(DTrainingCS, "belt")
histGroup(DTrainingCS, "[^(fore)]arm")
histGroup(DTrainingCS, "dumbbell")
histGroup(DTrainingCS, "forearm")
```

##Train a prediction model

Using random forest, the out of sample error should be small. The error will be estimated using the 40% probing sample. The error estimate that is less than 3% is consider good.

Set up the parallel clusters.

```{r}
library(parallel)
library(doParallel)
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)
```

Set the control parameters.

```{r}
ctrl <- trainControl(classProbs=TRUE,
                     savePredictions=TRUE,
                     allowParallel=TRUE)
```

Fir model over the tuning parameters.

```{r}
method <- "rf"
trainingModel <- train(classe ~ ., data=DTrainingCS, method=method)
```

Stop the clusters.

```{r}
stopCluster(cl)
```

##Evaluate the model on the training dataset

```{r}
trainingModel
hat <- predict(trainingModel, DTrainingCS)
confusionMatrix(hat, DTraining[, classe])
```

##Evaluate the model on the probing dataset

```{r}
hat <- predict(trainingModel, DProbingCS)
confusionMatrix(hat, DProbingCS[, classe])
```

##Display the final model

```{r}
varImp(trainingModel)
trainingModel$finalModel
```

The estimated error rate is less than **1%**.

Save training model object for later.

```{r}
save(trainingModel, file="trainingModel.RData")
```

##Predict on the test data

Load the training model.

```{r}
load(file="trainingModel.RData", verbose=TRUE)
```

Get the predictions and evaluate.

```{r}
DTestCS <- predict(preProc, DTest[, predCandidates, with=FALSE])
hat <- predict(trainingModel, DTestCS)
DTest <- cbind(hat , DTest)
subset(DTest, select=names(DTest)[grep("belt|[^(fore)]arm|dumbbell|forearm", names(DTest), invert=TRUE)])
```

